\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-16 doc/mlsys_paper/paper.tex
move mention of future optimizers to the conclusion
added TODO: moving the list of optimizers to section 2 would improve the flow and free up a bit of space
gain a bit of space by reducing interitem spacing in bullet lists
moved mention of GPUs and NVIDIA to the intro

{\tt ensmallen} is open-source software licensed under the 3-clause BSD
license~\cite{Rosen_2004_full}, 
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations.

@@

{\tt ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\argminright_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for {\tt
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.
%% CS: in the above sentence, it's not clear what is meant by "we must design ..."
%% CS: should it be "we provide ..." or "ensmallen provides ..." ?
%% CS: The overall intent of the sentence is not clear -- to my eyes
%% CS: it doesn't seem to gel well with its surrounding context.

@@

% These were a part of mlpack but not ensmallen.
  %\item {\bf Objective functions:} Neural Networks, Logistic regression,
  %    Matrix completion, Neighborhood Components Analysis, Regularized SVD,
  %    Reinforcement learning, Softmax regression, Sparse autoencoders,
  %    Sparse SVM
\end{enumerate}
\vspace*{-0.4em}

%% CS: "in development" is ambiguous here;
%% CS: is it in development by AI/ML researchers in general,
%% CS: or by ensmallen developers/contributors?
% In addition, many optimization algorithms are currently in development and will
% be released in the future.

%% CS: less ambiguous proclamation that has similar effect:
We aim to expand {\tt ensmallen} with further optimization techniques
as the need arises.  Since {\tt ensmallen} is open source,
external contributions to the codebase are welcome.

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item simulated annealing implementation in {\tt Optim.jl} with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} in the {\tt optim} package for GNU Octave~\cite{octave}
\end{itemize}
\vspace*{-0.3em}

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}-1: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{itemize}
\vspace*{-0.3em}

@@

GPU support for optimization is available either through the use of the NVBLAS
library~\cite{nvidia2015} or the forthcoming Bandicoot C++ GPU matrix library.
% Conrad: is it ok if we mention that here?  It's a really nice selling point
% and I am sure at least one reviewer will be thinking about GPUs.

>>>>

{\tt ensmallen} is open-source software licensed under the 3-clause BSD
license~\cite{Rosen_2004_full}, 
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations,
with optional interface to GPUs via NVBLAS~\cite{nvidia2015}.

@@

{\tt ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\argminright_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for {\tt
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.

%% TODO: moving the list of optimizers here would improve the flow 
%% TODO: and free up a bit of space

@@

% These were a part of mlpack but not ensmallen.
  %\item {\bf Objective functions:} Neural Networks, Logistic regression,
  %    Matrix completion, Neighborhood Components Analysis, Regularized SVD,
  %    Reinforcement learning, Softmax regression, Sparse autoencoders,
  %    Sparse SVM
\end{enumerate}

@@

\vspace*{-0.3em}
\begin{itemize}
\renewcommand{\itemsep}{-0.5ex}
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item simulated annealing implementation in {\tt Optim.jl} with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} in the {\tt optim} package for GNU Octave~\cite{octave}
\end{itemize}
\vspace*{-0.3em}

@@

\vspace*{-0.3em}
\begin{itemize}
\renewcommand{\itemsep}{-0.5ex}
  \item {\tt ensmallen}-1: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses {\tt optim} package for GNU Octave
\end{itemize}
\vspace*{-0.3em}

@@

The library provides many pre-built optimizers (including numerous variants
of SGD and Quasi-Newton optimizers).
We aim to expand the library with further optimization techniques
as the need arises.  Since {\tt ensmallen} is open source,
external contributions to the codebase are welcome.

% GPU support for optimization is available either through the use of the NVBLAS
% library~\cite{nvidia2015} or the forthcoming Bandicoot C++ GPU matrix library.
%% Conrad: is it ok if we mention that here?  It's a really nice selling point
%% and I am sure at least one reviewer will be thinking about GPUs.

%% CS: we don't have much room, so I moved the mention of NVBLAS to the intro.
%% CS: i think it flows better there.
%% CS: mentioning it here in the conclusion seems a bit left-field.
%% CS: we also don't have a good reference for Bandicoot yet
